\documentclass[12pt,a4paper]{article}
\begin{document}
\title{Rejoinder to: "Heteroscedastic Censored and Truncated Regression with crch"}
\maketitle
We thank the reviewer for his valuable suggestions and comments. The comments
and our corresponding replies can be found below:
%\textit{I enjoyed very much reading the manuscript, because it is well written
%and describes an interesting and relevant R package: “crch” for
%heteroscedastic censored and truncated regression. The package is also well
%prepared (e.g. the source code is well written and commented so that it is
%quite easy to read for humans) and well documented. In the following, I will
%give a few minor recommendations for improvement:}
\begin{enumerate}
  \item \textit{On page 1, last line, the term "log ($g(\sigma) =
  \log(\sigma)$)" is easily misunderstood as "$\log(g(\sigma) =
  \log(\sigma))$," which does not really make sense.  I suggest to write
  instead something like "logarithm, i.e. $g(\sigma) = \log(\sigma)$."} 
  \item \textit{In the last line of page 2 and in the following lines on page
  3, several tilde symbols (\textasciitilde) are missing. They can be obtained,
  e.g., by the \LaTeX command \textbackslash textasciitilde.} 
  \item \textit{On
  page 3, in the third from last paragraph, the last sentence should be
  terminated with a full stop (.).} 
  \item \textit{On page 3, in the last paragraph, I suggest to mention the
  software environment (R), for which the packages gamlss, gamlss.cens, and
  gamlss.tr are written for.}
\end{enumerate}
We have implemented these modifications as suggested by the reviewer.

\begin{enumerate}
  \setcounter{enumi}{4}
  \item \textit{I have some concern regarding the example with the two-part
  model: as far as I can see, this model does not take a into account a
  possible correlation between the error term of the logit (selection) model
  and the error term of the truncated (outcome) model, although it is quite
  likely that these two error terms are correlated. In order to allow for
  different processes that drive the occurrence of precipitation and the
  precipitation amount, a heteroscedastic(Heckman-type) sample selection model
  is probably more appropriate.}
\end{enumerate}

This is an interesting comment and it is indeed possible that correlations
between the error terms exist.  However, we think a (heteroscedastic) Heckman
sample selection model is not appropriate for our data since the non-zero
subset cannot be assumed to be normally distributed. A log-transformation which
is often applied in such cases, is also problematic since we then should also
transform the regressor (ensemble mean) that frequently takes on 0 values.

Alternatively, we tried to consider correlations in the error terms in a double
hurdle model (R-package mhurdle).  However, we did not find a significant
improvement with this model. 

We believe that comparing the tobit with a two-part model is a valid way to
test whether precipitation occurrence and amount can be assumed to be driven by
the same process. In our example the improved information criteria indicate
deficiencies in the tobit model specifications. Furthermore they also indicate
improved predictions by the two part model. Since in our weather forecast
example we are mainly interested in predictions, possibly incomplete model
assumptions (e.g., independence of error terms) might not be so relevant as
long as the predictions are fine.

We added a sentence: "In the censored model the occurrence of precipitation and
precipitation amount are assumed to be driven by the same process. To test this
assumption we compare the censored ..." to indicate that we are mainly
interested in testing the tobit model assumptions.
\end{document}
