---
title: "`procast`: Illustrate the package"
author: "Moritz N. Lang"
date: "March 2021"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
```{r include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.align="center") 

# install.packages("topmodels", repos="http://R-Forge.R-project.org")
# install.packages("crch", repos="http://R-Forge.R-project.org")
# install.packages("disttree", repos="http://R-Forge.R-project.org")

require("topmodels")
require("crch")
require("disttree")
```

## Procast
This is an demo file to illustrate the R package `procast`. Currently there are `procast()` S3 methods for the model classes `lm`, `crch`, and `disttree`. Both the latter in development status.  

After loading the necessary libraries, let's start by fitting some models:
```{r}
m1 <- lm(dist ~ speed, data = cars)
m2 <- crch(dist ~ speed | speed, left = 30, data = cars)
m3 <- distforest(dist ~ speed, data = cars, ntree = 10)
```

## Q-Q Plots for Quantile Residuals
```{r, fig.height = 2}
par(mar = c(4, 4, 1.5, 1.5), mfrow = c(1, 3))
qqrplot(m1, pch = 20, col = gray(0.4, alpha = 0.6))
qqrplot(m2, pch = 20, col = gray(0.4, alpha = 0.6))
qqrplot(m3, pch = 20, col = gray(0.4, alpha = 0.6))
```

## PIT histogram
Now we look at PIT histograms by calling `pithist()` which employs i.a. the functions `procast()`, `newresponse()`, and `qresiduals()`. Here solely an model dependent S3 method for `procast()` must be provided/written, all other R functions are ready to use.

```{r}
pit1 <- pithist(m1, plot = FALSE)
pit2 <- pithist(m2, plot = FALSE)
pit3 <- pithist(m3, plot = FALSE)
```

### PIT histograms with base R
There are two plotting styles available by setting the argument `style = c("histogram", "lines")`:

```{r}
plot(pit1, style = "histogram")
plot(pit1, style = "lines")
```

We can also combine several PIT histograms in a single plot:

```{r}
par(mar = c(4, 4, 1.5, 1.5))
plot(c(pit1, pit2, pit3), style = "histogram")
```

Or create first a PIT histogram with `style = "lines"` and than add other models:

```{R, fig.height = 3.5}
plot(pit1, style = "lines")
lines(c(pit2, pit3), col = c(2, 3))
```

### PIT histograms with `ggplot2`

```{r include=FALSE}
require("ggplot2")
```
There is also an `autoplot()` function available, which works for single models and also several models.
```{r, fig.width=4, fig.height = 3}
autoplot(pit1, style = "histogram")
```

When we plot several models using the `style = "lines"` we can specify the `colour` and `linetype` arguments, either a single value or a vector with the length of models:

```{r}
autoplot(c(pit1, pit2, pit3), style = "lines", 
         colour = c(1, 2, 3), linetype = 1)
```

We can also plot all in a single figure without using `facet_grid()`:

```{r, fig.height = 3}
autoplot(c(pit1, pit2, pit3), style = "lines", grid = FALSE, 
         colour = c(1, 2, 3), linetype = c(1, 2, 3))
```

## Discrete distributions `glm` model objects

```{r}
data("CrabSatellites", package = "countreg")
m_pois  <- glm(satellites ~ width + color, data = CrabSatellites, family = poisson)

CrabSatellites2 <- CrabSatellites[CrabSatellites$satellites <= 1, ]
m_binom <- glm(satellites ~ width + color, data = CrabSatellites2, family = binomial)

qqrplot(m_pois)
try(qqrplot(m_binom))
```


