---
title: 'CIs Comparison: `qqplotr` and `topmodels`'
output:
  html_document:
    df_print: paged
---

Start by loading the qqplotr package:

```{r echo=TRUE, message=FALSE}
require(qqplotr)
require(patchwork)
if(!require("devtools")) install.packages("devtools")
install_svn("svn://r-forge.r-project.org/svnroot/topmodels/pkg/topmodels",
            revision = "r1577", quiet = TRUE, upgrade = "never")
library("topmodels")
```
## `qqplotr` [vignette](https://aloy.github.io/qqplotr/)

Let’s start by simulating from a standard Normal distribution:

```{r}
set.seed(0)
smp <- data.frame(norm = rnorm(100))
```

Then, we use the provided stat_qq_* functions to construct a complete Q-Q plot with the points, reference line, and the confidence bands. As default, the standard Q-Q Normal plot with Normal confidence bands is constructed:

```{r}
gg <- ggplot(data = smp, mapping = aes(sample = norm)) +
    stat_qq_band() +
    stat_qq_line() +
    stat_qq_point() +
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles")
gg
```

As we can see, all the points lie within the confidence bands, which is expected for the given distribution.

As previously described in the Details section, three confidence bands constructs are available, which may be adjusted with the bandType parameter. Here, we may use the geom_qq_band instead of stat_qq_band, which permits a little more flexibility with the graphical parameters when constructing and visualizing different confidence bands.

```{r}
gg <- ggplot(data = smp, mapping = aes(sample = norm)) +
    geom_qq_band(bandType = "ks", mapping = aes(fill = "KS"), alpha = 0.5) +
    geom_qq_band(bandType = "ts", mapping = aes(fill = "TS"), alpha = 0.5) +
    geom_qq_band(bandType = "pointwise", mapping = aes(fill = "Normal"), alpha = 0.5) +
    geom_qq_band(bandType = "boot", mapping = aes(fill = "Bootstrap"), alpha = 0.5) +
    stat_qq_line() +
    stat_qq_point() +
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
    scale_fill_discrete("Bandtype")
gg
```

## Create `topmodel` object

As `qqplotr` does not work with model objects, I squeeze the same simulated data into a `topmodels` `qqrplot` object: 
```{r}
tmp_model <- lm(dist ~ speed, data = cars)
tmp_qqr <- qqrplot(tmp_model, plot = FALSE)

q2q <- function(y) qnorm(ppoints(length(y)))[order(order(y))]
compute_qthe <- function(x) apply(x, 2L, q2q)
qqr <- data.frame(
  observed = smp$norm, 
  expected = compute_qthe(smp),
 simint_observed_lwr = NA,
 simint_observed_upr = NA,
 simint_observed_expected = NA
)

attributes(qqr) <- attributes(tmp_qqr)
attr(qqr, "row.names") <- seq.int(1, length(smp$norm))
```

## Comparison of QQ Plots

### Point-wise CIs

Let's now compare the various implementations. Be aware, that `qqplotr` always plots polygons (with `geom_ribbon()`) for the various CI, whereas in `topmodels` the CI are visualized regardless the data using `curve()` etc. This allows us to plot also CI for areas without any data as typically seen in `wormplots`.

```{r}
qq1_qqplotr <- ggplot(data = smp, mapping = aes(sample = norm)) +
  geom_qq_band(
    bandType = "pointwise", 
    dparams = list(mean = 0, sd = 1),
    identity = FALSE, 
    qprobs = c(0.25, 0.75), 
    fill = "red", alpha = 0.5, show.legend = FALSE
  ) +
  stat_qq_line(
    dparams = list(mean = 0, sd = 1),
    identity = FALSE, 
    qprobs = c(0.25, 0.75),
    colour = "red"
  ) +
  stat_qq_point(
    dparams = list(mean = 0, sd = 1),
    colour = "black"
  ) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('qqrplot')

qq1_topmodels <- ggplot(qqr, aes(x = expected, y = observed)) + 
  geom_qqrplot_confint(
    type = "pointwise",
    identity = FALSE, 
    probs = c(0.25, 0.75),
    fill = "blue"
  ) + 
  geom_qqrplot_ref(
    identity = FALSE, 
    probs = c(0.25, 0.75),
    colour = "blue"
  ) + 
  geom_qqrplot(
    colour = "black"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('topmodels')

qq1_combined <- ggplot(data = smp) +
  geom_qq_band(
    mapping = aes(sample = norm), 
    bandType = "pointwise", 
    dparams = list(mean = 0, sd = 1),
    identity = FALSE, 
    qprobs = c(0.25, 0.75), 
    fill = "red", alpha = 0.5, show.legend = FALSE
  ) +
  stat_qq_line(
    dparams = list(mean = 0, sd = 1),
    mapping = aes(sample = norm), 
    identity = FALSE, 
    qprobs = c(0.25, 0.75),
    colour = "red"
  ) +
  stat_qq_point(
    dparams = list(mean = 0, sd = 1),
    mapping = aes(sample = norm), 
    colour = "red"
  ) +
  geom_qqrplot_confint(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    type = "pointwise",
    identity = FALSE, 
    probs = c(0.25, 0.75),
    fill = "blue"
  ) + 
  geom_qqrplot_ref(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    identity = FALSE, 
    probs = c(0.25, 0.75),
    colour = "blue"
  ) + 
  geom_qqrplot(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    colour = "blue"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('combined')

patchwork <- qq1_qqplotr + qq1_topmodels + qq1_combined
patchwork + plot_annotation(
  title = "QQ Plot (normal scale) with pointwise confidence intervals",
  subtitle = "without identity reference line"
)
```

```{r}
qq1b_qqplotr <- ggplot(data = smp, mapping = aes(sample = norm)) +
  geom_qq_band(
    bandType = "pointwise", 
    dparams = list(mean = 0, sd = 1),
    identity = TRUE, 
    fill = "red", alpha = 0.5, show.legend = FALSE
  ) +
  stat_qq_line(
    identity = TRUE, 
    dparams = list(mean = 0, sd = 1),
    colour = "black"
  ) +
  stat_qq_point(
    dparams = list(mean = 0, sd = 1),
    colour = "black"
  ) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('qqrplot')

qq1b_topmodels <- ggplot(qqr, aes(x = expected, y = observed)) + 
  geom_qqrplot_confint(
    type = "pointwise",
    identity = TRUE, 
    fill = "blue"
  ) + 
  geom_qqrplot_ref(
    identity = TRUE, 
    colour = "blue"
  ) + 
  geom_qqrplot(
    colour = "blue"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('topmodels')

qq1b_combined <- ggplot(data = smp) +
  geom_qq_band(
    mapping = aes(sample = norm), 
    bandType = "pointwise", 
    dparams = list(mean = 0, sd = 1),
    identity = TRUE, 
    fill = "red", alpha = 0.5, show.legend = FALSE
  ) +
  stat_qq_line(
    mapping = aes(sample = norm), 
    dparams = list(mean = 0, sd = 1),
    identity = TRUE, 
    colour = "red"
  ) +
  stat_qq_point(
    mapping = aes(sample = norm), 
    dparams = list(mean = 0, sd = 1),
    colour = "red"
  ) +
  geom_qqrplot_confint(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    type = "pointwise",
    identity = TRUE, 
    fill = "blue"
  ) + 
  geom_qqrplot_ref(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    identity = TRUE, 
    colour = "blue"
  ) + 
  geom_qqrplot(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    colour = "blue"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('combined')

patchwork <- qq1b_qqplotr + qq1b_topmodels + qq1b_combined
patchwork + plot_annotation(
  title = "QQ Plot (normal scale) with pointwise confidence intervals",
  subtitle = "with identity reference line"
)
```

### Simultaneous CIs

```{r}
qq2_qqplotr <- ggplot(data = smp, mapping = aes(sample = norm)) +
  geom_qq_band(
    bandType = "ks", 
    dparams = list(mean = 0, sd = 1),
    identity = FALSE, 
    qprobs = c(0.25, 0.75), 
    fill = "red", alpha = 0.5, show.legend = FALSE
  ) +
  stat_qq_line(
    dparams = list(mean = 0, sd = 1),
    identity = FALSE, 
    qprobs = c(0.25, 0.75),
    colour = "red"
  ) +
  stat_qq_point(
    dparams = list(mean = 0, sd = 1),
    colour = "black"
  ) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('qqrplot')

qq2_topmodels <- ggplot(qqr, aes(x = expected, y = observed)) + 
  geom_qqrplot_confint(
    type = "simultaneous",
    identity = FALSE, 
    probs = c(0.25, 0.75),
    fill = "blue"
  ) + 
  geom_qqrplot_ref(
    identity = FALSE, 
    probs = c(0.25, 0.75),
    colour = "blue"
  ) + 
  geom_qqrplot(
    colour = "black"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('topmodels')

qq2_combined <- ggplot(data = smp) +
  geom_qq_band(
    mapping = aes(sample = norm), 
    bandType = "ks", 
    dparams = list(mean = 0, sd = 1),
    identity = FALSE, 
    qprobs = c(0.25, 0.75), 
    fill = "red", alpha = 0.5, show.legend = FALSE
  ) +
  stat_qq_line(
    mapping = aes(sample = norm), 
    dparams = list(mean = 0, sd = 1),
    identity = FALSE, 
    qprobs = c(0.25, 0.75),
    colour = "red"
  ) +
  stat_qq_point(
    mapping = aes(sample = norm), 
    dparams = list(mean = 0, sd = 1),
    colour = "red"
  ) +
  geom_qqrplot_confint(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    type = "simultaneous",
    identity = FALSE, 
    probs = c(0.25, 0.75),
    fill = "blue"
  ) + 
  geom_qqrplot_ref(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    identity = FALSE, 
    probs = c(0.25, 0.75),
    colour = "blue"
  ) + 
  geom_qqrplot(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    colour = "blue"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('combined')

patchwork <- qq2_qqplotr + qq2_topmodels + qq2_combined
patchwork + plot_annotation(
  title = "QQ Plot (normal scale) with simultaneous confidence intervals",
  subtitle = "without identity reference line"
)
```

```{r}
qq2b_qqplotr <- ggplot(data = smp, mapping = aes(sample = norm)) +
  geom_qq_band(
    bandType = "ks", 
    dparams = list(mean = 0, sd = 1),
    identity = TRUE, 
    fill = "red", alpha = 0.5, show.legend = FALSE
  ) +
  stat_qq_line(
    dparams = list(mean = 0, sd = 1),
    identity = TRUE, 
    colour = "black"
  ) +
  stat_qq_point(
    dparams = list(mean = 0, sd = 1),
    colour = "black"
  ) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('qqrplot')

qq2b_topmodels <- ggplot(qqr, aes(x = expected, y = observed)) + 
  geom_qqrplot_confint(
    type = "simultaneous",
    identity = TRUE, 
    fill = "blue"
  ) + 
  geom_qqrplot_ref(
    identity = TRUE, 
    colour = "blue"
  ) + 
  geom_qqrplot(
    colour = "blue"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('topmodels')

qq2b_combined <- ggplot(data = smp) +
  geom_qq_band(
    mapping = aes(sample = norm), 
    dparams = list(mean = 0, sd = 1),
    bandType = "ks", 
    identity = TRUE, 
    fill = "red", alpha = 0.5, show.legend = FALSE
  ) +
  stat_qq_line(
    mapping = aes(sample = norm), 
    dparams = list(mean = 0, sd = 1),
    identity = TRUE, 
    colour = "red"
  ) +
  stat_qq_point(
    mapping = aes(sample = norm), 
    dparams = list(mean = 0, sd = 1),
    colour = "red"
  ) +
  geom_qqrplot_confint(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    type = "simultaneous",
    identity = TRUE, 
    fill = "blue"
  ) + 
  geom_qqrplot_ref(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    identity = TRUE, 
    colour = "blue"
  ) + 
  geom_qqrplot(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    colour = "blue"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('combined')

patchwork <- qq2b_qqplotr + qq2b_topmodels + qq2b_combined
patchwork + plot_annotation(
  title = "QQ Plot (normal scale) with simultaneous confidence intervals",
  subtitle = "with identity reference line"
)
```

### Tail-sensitive CIs

Here, we have the problem that in the current implementation we need the observed and expected data points which `geom_qqrplot_confint()` does not (yet) support. 

```{r}
qq3_qqplotr <- ggplot(data = smp, mapping = aes(sample = norm)) +
  geom_qq_band(
    bandType = "ts", 
    dparams = list(mean = 0, sd = 1),
    identity = TRUE, 
    fill = "red", alpha = 0.5, show.legend = FALSE
  ) +
  stat_qq_line(
    dparams = list(mean = 0, sd = 1),
    identity = TRUE, 
    colour = "red"
  ) +
  stat_qq_point(
    dparams = list(mean = 0, sd = 1),
    colour = "black"
  ) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('qqrplot')

tmp <- topmodels:::compute_qqrplot_confint(x = sort(qqr$observed), n = length(qqr$observed), type = "tail-sensitive")
qqr2 <- cbind(qqr, tmp)

qq3_topmodels <- ggplot(qqr2, aes(x = expected, y = observed)) + 
  geom_ribbon(
    mapping = aes(x = sort(expected), ymin = confint_lwr, ymax = confint_upr), 
    colour = NA, 
    fill = "blue", 
    alpha = 0.2
  ) +
  geom_qqrplot_ref(
    identity = TRUE, 
    colour = "blue"
  ) + 
  geom_qqrplot(
    colour = "black"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('topmodels')

qq3_combined <- ggplot(data = smp) +
  geom_qq_band(
    mapping = aes(sample = norm), 
    dparams = list(mean = 0, sd = 1),
    bandType = "ts", 
    identity = TRUE, 
    fill = "red", alpha = 0.5, show.legend = FALSE
  ) +
  stat_qq_line(
    mapping = aes(sample = norm), 
    dparams = list(mean = 0, sd = 1),
    identity = TRUE, 
    colour = "red"
  ) +
  stat_qq_point(
    mapping = aes(sample = norm), 
    dparams = list(mean = 0, sd = 1),
    colour = "red"
  ) +
  geom_ribbon(
    data = qqr2, 
    mapping = aes(x = sort(expected), ymin = confint_lwr, ymax = confint_upr), 
    colour = NA, 
    fill = "blue", 
    alpha = 0.2
  ) +
  geom_qqrplot_ref(
    data = qqr2, 
    mapping = aes(x = expected, y = observed),
    identity = TRUE,
    colour = "blue"
  ) + 
  geom_qqrplot(
    data = qqr, 
    mapping = aes(x = expected, y = observed),
    colour = "blue"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('combined')

patchwork <- qq3_qqplotr + qq3_topmodels + qq3_combined
patchwork + plot_annotation(
  title = "QQ Plot (normal scale) with tail-sensitive confidence intervals",
  subtitle = "with identity reference line"
)
```

Trying to intimate what `geom_qqrplot_confint()` currently does:

```{r}
x <- seq(min(qqr$expected), max(qqr$expected), length.out = 100)
tmp <- topmodels:::compute_qqrplot_confint(
  x = x,
  n = 100, 
  type = "tail-sensitive"
)
qqr3 <- cbind(qqr, tmp, x)

qq3b_topmodels <- ggplot(qqr3, aes(x = expected, y = observed)) + 
  geom_ribbon(
    mapping = aes(x = x, ymin = confint_lwr, ymax = confint_upr), 
    colour = NA, 
    fill = "blue", 
    alpha = 0.2
  ) +
  geom_qqrplot_ref(
    identity = TRUE, 
    colour = "blue"
  ) + 
  geom_qqrplot(
    colour = "black"
  ) + 
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  ggtitle('alternativ topmodels')

patchwork <- qq3_topmodels + qq3b_topmodels
patchwork + plot_annotation(
  title = "QQ Plot (normal scale) with tail-sensitive confidence intervals",
  subtitle = "with identity reference line"
)
```
Obviously this does not work. But shouldn't that still be working, e.g., for a perfect prediction where all points lie on the reference line?!

## Summary of comparison
Hence current issues are:

* Are the various implementations of the CIs correct.
* Why do the data points vary between `qqplotr` and `qqrplot`
    * Solved.
    * In `qqplotr`, per default distribution parameters are computed for the generation of theoretical quantiles.
    * This can be turned off by setting `dparams = list(mean = 0, sd = 1)`. 
    * Do we need something similar in `topmodels`?
* Why are there small differences in the tails for `type = "pointwise"`? 
    * Solved.
    * Added `slope` in numerator of the standard error (compare commit 1577).
* Can tail sensitive CIs also be implemented in order to work with `curve()` etc.
* Check if all CIs are valid for `scale = "uniform"`, here `qqplotr` has solely a single CI implemented based on a bootstrap.
* Check if all CIs are valid for worm plots (`detrended = TRUE`).
