---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file and run rmarkdown::render("README.Rmd") -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/README-"
)
options(digits = 4)
```

## Infrastructure for Inference and Forecasting in Probabilistic Models

Unified infrastructure for probabilistic models and distributional regressions:
Computations of probabilities, densities, scores, and Hessians.
Probabilistic forecasting. Diagnostic graphics such as rootograms, PIT histograms,
reliagrams (reliability diagrams), and (randomized) quantile residual Q-Q plots.

Modular object-oriented implementation with support for many model objects, including
`lm`, `glm`, `crch`, `disttree`, and more to come.

```{r topmodels, echo=FALSE, results="hide"}
library("topmodels")
library("ggplot2")
data("CrabSatellites", package = "countreg")
CrabSatellites2 <- CrabSatellites[CrabSatellites$satellites <= 1, ]

m1 <- glm(satellites ~ width + color, data = CrabSatellites, family = poisson)

theme_set(theme_minimal())
svg(filename = paste0(file.path(getwd(), "man", "figures"), "/README-topmodels.svg"),
  width = 9, height = 7.5)

topmodels(m1, plot = "ggplot2", which = 1:4, single_page = TRUE)

dev.off()
```

<img alt="Probabilistic model evaluation with topmodels" src="man/figures/README-topmodels.svg" style="border:10px solid transparent">
