---
title: "Graphical Evaluation: Methodology"
author: "Moritz Lang, Achim Zeileis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{Graphical Model Assessment under Different Model Misspecifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{topmodels,crch,MASS,rmutil,sn}
  %\VignetteKeywords{FIXME}
  %\VignettePackage{topmodels}
---

```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

```{r preliminaries, echo = FALSE, message = FALSE}
library("topmodels")
library("crch")
library("MASS")
library("rmutil")  # for laplace dist
library("sn") # for skewed N dist
options(digits = 4)
knitr::opts_chunk$set(fig.aling = "center")

make_residuals_plot <- function(object, breaks = "Sturges", main = "") {
  name <- deparse(substitute(object))

  set.seed(1)
  hist(qresiduals(object), breaks = breaks, freq = FALSE,
    main = main,
    xlab = "Q-Q residuals", xlim = c(-5, 5), ylim = c(0, 1)
  )
  curve(dnorm, from = -5, to = 5, col = 2, lwd = 2, add = TRUE)
  legend("topright", "std. norm", lty = 1, col = 2, lwd = 2, bty = "n")
  box()
}
```

```{r dgp, echo = FALSE}
set.seed(0)

## regressors
d <- data.frame(
  x = runif(500, -1, 1),
  z = runif(500, -1, 1)
)

d <- transform(d,
  y_norm = rnorm(500),
  y_rskew = sn::rsn(500, xi = 0, omega = 1, alpha = 5, tau = 0),
  y_lskew = sn::rsn(500, xi = 0, omega = 1, alpha = -5, tau = 0),

  y_htail = rt(500, df = 4),
  y_ltail = c(rnorm(100), runif(400, min = -1.5, max = 1.5)),

  y_odisp = rpois(500, lambda = exp(0 + 2 * x + rnorm(500, mean = 0, sd = 1.5))),
  y_udisp = round(rnorm(n = 500, mean = exp(0 + 2 * x), sd = 0.001))

)
```

```{r fit, echo = FALSE}
m_norm <- lm(y_norm ~ 1, data = d)
m_rskew <- lm(y_rskew ~ 1, data = d)
m_lskew <- lm(y_lskew ~ 1, data = d)
m_htail <- lm(y_htail ~ 1, data = d)
m_ltail <- lm(y_ltail ~ 1, data = d)

m_odisp <- glm(y_odisp ~ x, family = "poisson", data = d)
m_udisp <- glm(y_udisp ~ x, family = "poisson", data = d)
```

## Rootogram

The rootogram is a graphical tool for assessing the goodness of fit of a
parametric univariate distributional model, with estimated parameters
$\hat{\boldsymbol{\theta}}_{i} = (\hat{\theta}_{i1}, \ldots,
\hat{\theta}_{iK})^\top$ and $\mathcal{f}( \cdot )$ desribing the density or
probability mass function. 
Rootograms evaluate graphically whether observed frequencies
$\text{obs}_j$ match the expected frequencies $\text{exp}_j$ by plotting 
histogram-like rectangles or bars for the observed frequencies 
and a curve for the fitted frequencies, both on a square-root scale. In the form
presented here, it was implemented by @Kleiber+Zeileis:2016 building on work of @Tukey:1977. 

In the most general form, given an observational vector of a random variable
$y_i (i = 1, \ldots, n)$ which is divided into subsets by a set of breakpoints $b_0,
b_1, b_2, \dots~$, the observed and expected frequencies are given by

$$\text{obs}_j = \sum_{i=1}^{n}w_{i} I(y_i \in (b_j, b_{j+1}]),$$
$$\text{exp}_j = \sum_{i=1}^{n}w\{\mathbf{\mathcal{F}}(b_{j+1} |
  \hat{\boldsymbol{\theta}}_{i}) - \mathbf{\mathcal{F}}(b_{j} | \hat{\boldsymbol{\theta}}_{i})\},$$

with $\mathbf{\mathcal{F}}( \cdot )$ being the CDF of the modeled
distributional model $\mathcal{f}( \cdot )$ and $w_i$ being optional
observation-specific weights. Whereby, the weights are typically needed either
for survey data or for situations with model-based weights
[@Kleiber+Zeileis:2016].

For a discrete variable $y_i$, the observed and expected frequencies can be
simplified and are given for each integer $j$ by

$$\text{obs}_j = \sum_{i=1}^{n}I(y_i - j),$$
$$\text{exp}_j = \sum_{i=1}^{n}\mathcal{f}(j | \hat{\boldsymbol{\theta}}_{i}),$$

with the indicator variable $I( \cdot )$ [@Kleiber+Zeileis:2016]. As rootograms are
best known for count data, the latter form is quite common. 

Different styles of rootograms have been proposed as extensively discussed in
@Kleiber+Zeileis:2016. As default, they propose a so called "hanging"
rootogram, which aligns all deviations along the horizontal axis, as the
rectangles are drawn from $\sqrt{exp_j}$ to $\sqrt{exp_j} - \sqrt{obs_j}$, so
that they "hang" from the curve with the expected frequencies $\sqrt{exp_j}$.

The concept of comparing observed and expected frequencies graphically was also
introduced in the seminal work on assessing calibration and sharpness for a
predictive probalilty model by @Gneiting+Balabdaoui+Raftery:2007 and, building
on this, applied to count data by @Czado+Gneiting+Held:2009.  However, since in
both cases either the deviations or the expected and observed frequencies are
presented only as lines connecting the respective frequencies, deviations are
more difficult to detect compared to the rootograms introduced by @Tukey:1977
and further enhanced by @Kleiber+Zeileis:2016.


## PIT Histogram

To check the calibration of a regression model, @Dawid:1984 proposed the use of
the probability integral transform (PIT) which is simply the predictive
cumulative distribution function (CDF) evaluated at the observations. For a
continuous random variable $y_i (i = 1, \ldots, n)$, it is defined as 

$$u_i = \mathbf{\mathcal{F}}(y_i | \, \hat{\boldsymbol{\theta}}_i)$$

 where $\mathbf{\mathcal{F}}( \cdot )$ denotes the CDF of the modeled
distribution $\mathcal{f}( \cdot )$ with estimated parameters
$\hat{\boldsymbol{\theta}}_{i} = (\hat{\theta}_{i1}, \ldots,
\hat{\theta}_{iK})^\top$. If the estimated model is a good approximation to the
true data generating process, the observation will be drawn from the predictive
distribution and the PIT values $u_i$ are approximately uniformly distributed
on $[0, 1]$. Plotting the histogram of the PIT values and checking for
uniformity is therefore a common empirical way of checking for calibration
[@Diebold+Gunther+Tay:1998; @Gneiting+Balabdaoui+Raftery:2007].  Whereas,
deviations from uniformity point to underlying forecast errors and model
deficiencies: U-shaped histograms refer to underdispersed predictive
distributions, inverted U-shaped histograms to overdispersion, and skewed
histograms suggest that central tendencies must be biased
[@Czado+Gneiting+Held:2009].

When considering discrete response distributions or distributions with a
discrete component, e.g., in case of censoring, for a random discrete variable
$y_i$ the PIT $u_i$ is generated as a random draw from the interval
$\mathbf{\mathcal{F}}(y_i | \, \hat{\boldsymbol{\theta}}_i)$. Even if this
leads to some randomness in the graphical representation of PIT values, in most
cases the impact on the graphical evaluation when repeating the calculations
(i.e. drawing new values $u_i$) is rather small. (FIXME: Compare with
@Czado+Gneiting+Held:2009 and their references to random PITs).


## Q-Q Residuals Plot

Quantile residuals are simply the inverse cumulative distribution function of a
standard normal distribution $\Phi^{-1}$ evaluated at values PIT values $u_i (i
= 1, \ldots, n)$, hence, they can be defined as 

$$\hat{r}_i = \Phi^{-1}(\mathbf{\mathcal{F}}(y_i | \, \hat{\boldsymbol{\theta}}_{i})) = 
  \Phi^{-1}(u_i),$$ 

where $\mathbf{\mathcal{F}}( \cdot )$ again denotes the cumulative distribution
function (CDF) of the modeled distribution $\mathcal{f}( \cdot )$ with
estimated parameters $\hat{\boldsymbol{\theta}}_{i} = (\hat{\theta}_{i1},
\ldots, \hat{\theta}_{iK})^\top$ [@Dunn+Smyth:1996]. As before, for discrete
or partly discrete responses, the approach includes some randomization to
achieve continuous $u_i$ values; quantile residuals are therefore often
referred to as randomized quantile residuals in the literature
[@Dunn+Smyth:1996]. 

In case of a correct model fit, the values $u_i$ are uniformly distributed on
the unit interval and the Q-Q residuals should at least approximately be
standard normally distributed. Hence, to check for normality, quantile
residuals can be graphically compared to theoretical quantiles of the standard
normal distribution, where strong deviations from the bisecting line indicate a
misspecified model fit. 

Mathematically, Q-Q plot consists of the tuples

$$(z_{(1)},  \hat{r}_{(1)}), \ldots,  (z_{(n)},  \hat{r}_{(n)}),$$

where $\hat{r}_{(i)}$ denotes the $i$th order statistic of the quantile
residuals, so that $\hat{r}_{(1)} \leq \hat{r}_{(2)} \cdot \leq \hat{r}_{(n)}$,
and $z_{(i)}$ is the ordered statistics from the respective standard normal
quantiles $\Phi^{-1}( p_i)$ evaluated at the cumulative proportion $p_i$. This
graphical evaluation is well known as normal probability plot or normal Q-Q
plot [@Hoaglin:2006]. Due to the transformation of the PIT values $u_i$ to the
normal scale, their extreme values are more widely spread, so that normal Q-Q
diagrams are better suited than, for example, PIT histograms to detect
violations of the distribution assumption within its tails. An additional
possible advantage of Q-Q plots is that they avoid the necessity of defining
breakpoints as typically needed for histogram style evaluations
[@Klein+Kneib+Lang+Sohn:2015].

But Q-Q plots can also be applied to check if residuals follow any other known
distribution, by employing any other inverse cumulative distribution function
of interest instead of $\Phi^{-1}$ in the computation and comparing the
quantile residuals $\hat{r}_i$ to the respective theoretical quantiles. This is
called than a theoretical quantile-quantile plot or Q-Q plot for short
[@Friendly:1991].


## Worm plot

As in Q-Q plots, small too medium deviations can be quite hard to detect,
untilting the plot by subtracting the theoretical quantiles, makes detecting
pattern of departure from a now horizontal line much easier. Mathematically,
therefore, the tuples in the plot are

$$(z_{(1)},  \hat{r}_{(1)} - z_{(1)}), \ldots,  (z_{(n)},  \hat{r}_{(n)} - z_{(n)}),$$

where as before, where $\hat{r}_{(i)}$ denotes the order statistic of the
empirical quantile residuals and $z_{(i)}$ the ordered statistics of the
respective standard normal quantiles.  This so-called de-trended Q-Q plot
[@Friendly:1991] is best known by the application of @Buuren+Fredriks:2001, and
is therefore usually referred to as worm plot according to their naming.

## Summary plot
```{r plot-qq, echo = FALSE, results = "hide"}
grDevices::svg(file = "figures/graph_eval_meth_qq.svg", width = 14, height = 14)
par(mfrow = c(6, 7), bg = "#f7f7f7")
make_residuals_plot(m_norm, breaks = 20, main = "Well calibrated")
make_residuals_plot(m_rskew, breaks = 20, main = "Too skew to the left")
make_residuals_plot(m_lskew, breaks = 20, main = "Too skew to the right")
make_residuals_plot(m_htail, breaks = 20, main = "Tails too light")
make_residuals_plot(m_ltail, breaks = 20, main = "Tails too heavy")
make_residuals_plot(m_odisp, main = "Underdispersed")
make_residuals_plot(m_udisp, main = "Overdispersed")
qqrplot(m_norm, main = "Q-Q residuals plot")
qqrplot(m_rskew, main = "Q-Q resiudals plot")
qqrplot(m_lskew, main = "Q-Q resiudals plot")
qqrplot(m_htail, main = "Q-Q resiudals plot")
qqrplot(m_ltail, main = "Q-Q resiudals plot")
qqrplot(m_odisp, main = "Q-Q resiudals plot")
qqrplot(m_udisp, main = "Q-Q resiudals plot")
#qqrplot(m_norm, trafo = identity, main = "Uniform Q-Q residuals plot")
#qqrplot(m_rskew, trafo = identity, main = "Uniform Q-Q resiudals plot")
#qqrplot(m_lskew, trafo = identity, main = "Uniform Q-Q resiudals plot")
#qqrplot(m_htail, trafo = identity, main = "Uniform Q-Q resiudals plot")
#qqrplot(m_ltail, trafo = identity, main = "Uniform Q-Q resiudals plot")
#qqrplot(m_odisp, trafo = identity, main = "Uniform Q-Q resiudals plot")
#qqrplot(m_udisp, trafo = identity, main = "Uniform Q-Q resiudals plot")
wormplot(m_norm, main = "Worm plot")
wormplot(m_rskew, main = "Worm plot")
wormplot(m_lskew, main = "Worm plot")
wormplot(m_htail, main = "Worm plot")
wormplot(m_ltail, main = "Worm plot")
wormplot(m_odisp, main = "Worm plot")
wormplot(m_udisp, main = "Worm plot")
pithist(m_norm, main = "PIT histogram")
pithist(m_rskew, main = "PIT histogram")
pithist(m_lskew, main = "PIT histogram")
pithist(m_htail, main = "PIT histogram") 
pithist(m_ltail, main = "PIT histogram") 
pithist(m_odisp, main = "PIT histogram")
pithist(m_udisp, main = "PIT histogram")
pithist(m_norm, main = "Customized PIT histogram",
  breaks = c(0, 0.005, seq(0,1,length.out = 22)[-c(1,22)], 0.995, 1)) 
pithist(m_rskew, main = "Customized PIT histogram",
  breaks = c(0, 0.005, seq(0,1,length.out = 22)[-c(1,22)], 0.995, 1)) 
pithist(m_lskew, main = "Customized PIT histogram",
  breaks = c(0, 0.005, seq(0,1,length.out = 22)[-c(1,22)], 0.995, 1)) 
pithist(m_htail, main = "Customized PIT histogram", 
  breaks = c(0, 0.005, seq(0,1,length.out = 22)[-c(1,22)], 0.995, 1)) 
pithist(m_ltail, main = "Customized PIT histogram", 
  breaks = c(0, 0.005, seq(0,1,length.out = 22)[-c(1,22)], 0.995, 1))
pithist(m_odisp, main = "Customized PIT histogram",
  breaks = c(0, 0.005, seq(0,1,length.out = 22)[-c(1,22)], 0.995, 1)) 
pithist(m_udisp, main = "Customized PIT histogram",
  breaks = c(0, 0.005, seq(0,1,length.out = 22)[-c(1,22)], 0.995, 1)) 
rootogram(m_norm, main = "Rootogram")
rootogram(m_rskew, main = "Rootogram")
rootogram(m_lskew, main = "Rootogram")
rootogram(m_htail, main = "Rootogram")
rootogram(m_ltail, main = "Rootogram")
rootogram(m_odisp, main = "Rootogram")
rootogram(m_udisp, main = "Rootogram")
par(mfrow = c(1, 1))
dev.off()
```

![](figures/graph_eval_meth_qq.svg){width=100%}

Predictive distribution | well calibrated | too skew to the left | too skew to the right | tails too light | tails too heavy | underdispersed (underestimated variance) | overdispersed (overestimated variance)
--- | --- | --- | --- | --- | --- | --- | ---
**Resdiuals** | normal | right skewed | left skewed | heavy tailed | light tailed | overdispersed | underdispersed
**Q-Q plot** | bisecting line | positive curvature (bends above the line at both ends) | negative curvature (bends down at both ends) | reverse S-shape (dips below the line at the low end and rises above it at the high end) | S-shape | crossing qqline from below (?) | crossing qqline from above ?
**Worm plot** | horizontal line | U-shape | inverse U-shape | S-shape on the left bent down | S-shape on the left bent up | positive slope | negative slope
**PIT histogram** | uniform | skewed | skewed | superimposed U-shape | superimposed inverse U-shape" | U-shape | inverse U-shape 
**Rootogram** | no deviations | ? | ? | wave-like (underfitting in the tails and the center) | ? | ? | ?
**Interpretation** | no misspecifications | ? | ? |  values more extrem as expected |  values less extrem as expected | ? | ?

## References

