---
title: "Illustration: Precipitation Forecasts in Innsbruck"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{Illustration: Precipitation Forecasts in Innsbruck}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{topmodels,crch,ggplot2}
  %\VignetteKeywords{FIXME}
  %\VignettePackage{topmodels}
---

```{r preliminaries, echo = FALSE, message = FALSE}
library("topmodels")
library("crch")
library("ggplot2")
knitr::opts_chunk$set(fig.align = "center")
```

The second example models 3 day-accumulated precipitation sums using a Logistic
distribution censored at zero accounting for non-negative precipitation sums.
The use case builds heavily on the vignette ["Heteroscedastic Censored and Truncated
Regression with crch"](https://cran.r-project.org/web/packages/crch/vignettes/crch.pdf) given in
[`crch`](https://CRAN.R-project.org/package=crch) by
@Messner+Mayr+Zeileis:2016. 

## Precipitation forecasts

This use-case discusses a weather forecast example application of censored
regression models. The example is taken from the vignette ["Heteroscedastic
Censored and Truncated Regression with
crch"](https://cran.r-project.org/web/packages/crch/vignettes/crch.pdf)
provided for the package [`crch`](https://CRAN.R-project.org/package=crch) by
@Messner+Mayr+Zeileis:2016. 

Weather forecasts are usually based on numerical weather prediction (NWP)
models, which take the current state of the atmosphere and calculate future
weather by numerically simulating the main atmospheric processes. However, due
to uncertain initial conditions and unknown or unresolved processes, these
numerical predictions are always subject to errors.  To estimate these errors,
many weather centers produce what are called ensemble forecasts: multiple NWP
runs that use different initial conditions and model formulations.
Unfortunately, these ensemble forecasts cannot account for all sources of
error, so they are often still biased and uncalibrated. Therefore, they are
often calibrated and corrected for systematic errors through statistical
post-processing.

One popular post-processing method is heteroscedastic linear regression where
the ensemble mean is used as regressor for the location and the ensemble
standard deviation or variance is used as regressor for the scale
[@Gneiting+Raftery+Westveld+Goldman:2005]. The following example applies
heteroscedastic censored regression with a logistic distribution assumption to
precipitation data in Innsbruck (Austria).

The `RainIbk` data set contains observed 3 day-accumulated precipitation
amounts (`rain`) and the corresponding 11 member ensemble forecasts of total
accumulated precipitation amount between 5 and 8 days in advance (`rainfc.1`,
`rainfc.2`, ... `rainfc.11`).  In previous studies it has been shown that it is
of advantage to model the square root of precipitation rather than
precipitation itself. Thus all precipitation amounts are square rooted before
ensemble mean and standard deviation are derived. Furthermore, events with no
variation in the ensemble are omitted:


```{r crch data}
data("RainIbk", package = "crch")

RainIbk <- sqrt(RainIbk)
RainIbk$ensmean <- apply(RainIbk[,grep('^rainfc',names(RainIbk))], 1, mean)
RainIbk$enssd <- apply(RainIbk[,grep('^rainfc',names(RainIbk))], 1, sd)
RainIbk <- subset(RainIbk, enssd > 0)
```

## From linear regression to distributional regression models

For comparison we fit a homoscedastic linear regression model using the least
squares approach for `rain` with `ensmean` as regressor. As a more appropriate
model for `rain` accounting for non-negative precipitation sums, we fit a
hetereoscedastic censored model assuming an underlying logistic response
distribution with `ensmean` as regressor for the location and `log(enssd)` as
regressor for the scale:

```{r crch models}
## linear model
m_lm <- lm(rain ~ ensmean, data = RainIbk)

## heteroscedastic censored regression with a logistic distribution assumption
m_hclog <- crch::crch(rain ~ ensmean | log(enssd), data = RainIbk, left = 0,
  dist = "logistic")
```

In the scatterplot of `rain` against `ensmean` it can be seen, that
precipitation is clearly non-negative with many zero observations. Thus the
censored regression model `m_hclog` is more suitable than the linear regression
model to estimate the underlying relationship:

```{r}
plot(rain ~ ensmean, data = RainIbk, pch = 19, col = gray(0, alpha = 0.2))
abline(coef(m_lm)[1:2], col = 3, lwd = 2)
abline(coef(m_hclog)[1:2], col = 4, lwd = 2)

legend("topright", lwd = c(2, 1), lty = c(1, 1), col = c(3, 4), 
  c("m_lm", "m_hclog"), bty = "n")
```


## Model evaluation: Marginal calibration

The significant underfitting of zero observations is also very apparent in the
_rootogramm_ of the linear model, as it does not correctly account for the
point mass at zero. Additionally, a weak wavelike pattern indicates a 
slight overfitting of precipitation 
sums between zero and 5mm and an underfitting of precipitation sums above. 
In contrast, the censored logistic regression `m_hclog`
provides a pretty good marginal fit, where the expected squared frequencies
closely match the squared observed frequencies:

```{r}
r1 <- rootogram(m_lm, plot = FALSE)
r2 <- rootogram(m_hclog, plot = FALSE)

ggplot2::autoplot(c(r1, r2))

```


## Model evaluation: Probabilistic calibration

The _PIT histogram_ of the linear model `m_lm` is skewed, indicating a misfit
in terms of the probabilistic calibration, whereas the _PIT histogram_ of the 
linear model is rather uniformly distributed.

```{r}
p1 <- pithist(m_lm, plot = FALSE)
p2 <- pithist(m_hclog, plot = FALSE)

ggplot2::autoplot(c(p1, p2))
```

For a better comparison of the two model fits, alternatively a line-style
_PIT histogram_ can also be shown:

```{r}
ggplot2::autoplot(c(p1, p2), colour = c(3, 4), style = "line", single_graph = TRUE, confint_col = 1, confint = "line", ref = FALSE)
```

To further analyse the misfit of the linear model in terms of probabilistic
calibration, we transform the PIT residuals to the normal scale and increase
the number of breaks in the _PIT histogram_. Transforming to the normal scale
spreads the values at the tails of the distribution further apart, and
increasing the breaks prevents potential masking of small scale patterns.

We can see that the distribution of PIT residuals is right skewed with more
values between -2 and 0, but less values between 0 and 2 compared to the
standard normal distribution:

```{r}
pithist(m_lm, breaks = 50, trafo = qnorm, xlim = c(-3, 3))
```

This is also reflected in the pattern of the _Q-Q plot_ of the linear model:
Due to the high frequency in values between -2 and 0, the observed quantile
residuals increase slower in this region relative to the standard normal
quantiles.  However, the low frequency of quantile residuals above zero leads
to an higher increase in quantile residuals compared to the standard normal
quantiles. 

In summary, the _Q-Q plot_ of the linear model has a positive curvature,
bending up at both ends of the distributional tails. The censored logistic
regression `m_hclog` follows more or less the reference line indicating
normally distributed quantile residuals:

```{r}
q1 <- qqrplot(m_lm, plot = FALSE)
q2 <- qqrplot(m_hclog, plot = FALSE)
ggplot2::autoplot(c(q1, q2), colour = c(3, 4), alpha = 0.8, single_graph = TRUE, simint = FALSE, confint ="line")
```

Consistent with the pattern of right-skewed PIT residuals corresponding to a
"too left-skewed" fitted distribution, a U-shaped pattern is evident in the
_worm plot_ of the linear model; the censored logistic is more or less flat
around zero.

```{r}
w1 <- wormplot(m_lm, plot = FALSE)
w2 <- wormplot(m_hclog, plot = FALSE)
ggplot2::autoplot(c(w1, w2), colour = c(3, 4), alpha = 0.8, single_graph = TRUE, simint = FALSE, confint ="line")
```

## References


