---
title: "Probabilistic Forecasting Infrastructure"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{Probabilistic Forecasting Infrastructure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{topmodels,distributions3,crch}
  %\VignetteKeywords{probabilistic forecasting, scoring rules, probability distributions, quantiles, simulation}
  %\VignettePackage{topmodels}
---

```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

```{r preliminaries, echo = FALSE, message = FALSE}
library("topmodels")
library("distributions3")
library("crch")
options(digits = 4)
knitr::opts_chunk$set(fig.align = "center")
```

## Overview

The functionality provided by the `topmodels` package can be broadly placed into three groups:

1. _Numeric quantities:_ Functions which, based on fitted model objects
  (like `lm`, `glm`, etc.), compute quantities of interest, such as predicted probabilities, quantiles,
  residuals, etc.
2. _Visualizations:_ Functions which can help assess goodness of fit for fitted model objects, leveraging
  the numeric quantities from above.
3. _Under the hood:_ Functions which can extract/predict probability distributions as proper S3 objects
  and provide standard methods for working with these distributions.
  
The S3 framework for probability distributions (3. above) is actually set up in the `distributions3` package
that `topmodels` builds upon. Actually, all numeric and graphical functions (1. and 2. above) have "smart"
default methods. This means that _if necessary_ methods can be defined for them but all the default methods
work out of the box if the `distributions3` functionality from 3. is provided. The table below shows
all of the functions that become directly available when interfacing the `distributions3` infrastructure.


| Function             | Description                                                                    |
|:---------------------|:-------------------------------------------------------------------------------|
| _Numeric quantities_ |                                                                                |
| `procast()`          | Probabilistic forecats (probabilities, quantiles, etc.) based on model objects |
| `proscore()`         | Evaluate scoring rules for procasts                                            |
| `pitresiduals()`     | Probability integral transform (PIT) residuals                                 |
| `qresiduals()`       | (Randomized) quantile residuals                                                |
| _Visualizations_     |                                                                                |
| `pithist()`          | PIT histograms                                                                 |
| `qqrplot()`          | Q-Q plots for quantile residuals                                               |
| `wormplot()`         | Worm plots for quantile residuals                                              |
| `rootogram()`        | Rootograms of observed and fitted frequencies                                  |
| `reliagram()`        | (Extended) reliability diagrams                                                |
| _Under the hood_     |                                                                                |
| `prodist()`          | Fitted and predicated probability distributions based on model objects         |
| `Normal()`,          | Probability distribution objects                                               |
| `Poisson()`, ...     | With methods `pdf()`, `cdf()`, `quantile()`, `mean()`, ...                     |


Thus, to connect a new class of models to the `topmodels` tools the following building blocks need to be
provided:

a. A class with all necessary methods for the probability distribution of the response variable.
b. A `prodist()` method for the model object, which typically first predicts the parameters and then
  sets up the distribution object from a.
  
Below we illustrate both steps by setting up a so-called "Tobit" model [@Tobin:1958]. This is a model with a normally-distributed
response, left-censored at zero, which can in R be fitted with the `crch()` function [@Messner+Mayr+Zeileis:2016] from the package
of the same name (among other packages). For illustration we first set up a new `Tobit()` distributions object
(which is a special case of the `CensoredNormal()` class provided in `crch`). Subsequently, we add a
`prodist()` method for `crch` objects.



## Adding a new distribution

To illustrate how to set up a `distributions3` constructor function, we employ the "Tobit" model, i.e.,
a normal distribution, left-censored at zero. Note that in practice we could just use the `CensoredNormal()`
distribution provided by the `crch()` package (with arbitrary left and/or right censoring) as well as the
underlying `dcnorm()`, `pcnorm()`, `qcnorm()`, and `rcnorm()` functions. However, here we avoid doing
so and set up the `Tobit()` class constructor and accompanying methods from scratch, just for illustration.


### Class constructor

First, the constructor (or generator) function for the distribution object should set up a data frame,
containing the distribution's parameters, with a custom class `"Tobit"` inheriting from `"distribution"`.
In case of the Tobit distribution, the parameters are the mean `mu` $= \mu$ and standard deviation
`sigma` $= \sigma$ of the underlying uncensored normal distribution.

```{r}
Tobit <- function(mu = 0, sigma = 1) {
  n <- c(length(mu), length(sigma))
  stopifnot("parameter lengths do not match (only scalars are allowed to be recycled)" =
    all(n %in% c(1L, max(n))))
  d <- data.frame(mu = mu, sigma = sigma)
  class(d) <- c("Tobit", "distribution")
  return(d)
}
```

This is already sufficient for setting up a vector `Y` containing three different Tobit distributions:

```{r}
Y <- Tobit(mu = 1:3, sigma = c(1, 1, 4))
Y
```

Printing, subsetting, and some coercion functions already work due to suitable, flexible methods for
`"distribution"` objects in general.

```{r}
length(Y)
Y[-2]
as.matrix(Y)
```


### Methods

Having constructed a distributions object like the `"Tobit"` object `Y` above, the next step is to
perform standard tasks on it, such as computing densities, probabilities, moments, etc. While in
base R the familiar functions of type `ddist()` (density), `pdist()` (cumulative probability),
`qdist()` (quantile), and `rdist()` (random numbers) are typically used (e.g., where `dist` = `norm`
or `pois` etc.), the `distributions3` employs an object-oriented approach. Thus, rather than having
the specification of the distribution as part of the function name, it is captured in the class of
the object. Based on that suitable generic functions like `pdf()` (density), `cdf()` (cumulative
probability), etc. can be provided for each distribution class. See the table below for an overview
of generic functions and their purpose.

| Function             | Package          | Description                                                                          |
|:---------------------|:-----------------|:-------------------------------------------------------------------------------------|
| _Distributional_     |                  |                                                                                      |
| `pdf()`              | `distributions3` | Probability density function (or probability mass function, typically via `ddist()`) |
| `log_pdf()`          | `distributions3` | Log-density (or log-likelihood, typically via `ddist(..., log = TRUE)`)              |
| `cdf()`              | `distributions3` | Cumulative distribution function (typically via `pdist()`)                           |
| `quantile()`         | `stats`          | Compute quantiles (typically via `qdist()`)                                          |
| `random()`           | `distributions3` | Simulate random samples (typically via `rdist()`)                                    |
| `crps()`             | `scoringRules`   | (Continuous) ranked probability scored (typically via `crps_dist()`)                 |
| _Moments_            |                  |                                                                                      |
| `mean()`             | `base`           | Expectation                                                                          |
| `variance()`         | `distributions3` | Variance                                                                             |
| `skewness()`         | `distributions3` | Skewness                                                                             |
| `kurtosis()`         | `distributions3` | Excess kurtosis                                                                      |
| _Other_              |                  |                                                                                      |
| `support()`          | `distributions3` | Maximum and minimum of the support of the probability distribution                   |

In the following we just show how the `mean()` and the `cdf()` method can be set up and applied. Some further details
are provided at the end of this section, see also the underlying source code in the packages.

First, the extractor functions for moments of the distribution can be typically be defined as functions of the list of
parameters stored in the object:

```{r}
mean.Tobit <- function(x, ...) {
  m <- x$mu * pnorm(x$mu/x$sigma) + x$sigma * dnorm(x$mu/x$sigma)
  setNames(m, names(x))
}
```

In a Tobit distribution the expectation is that of the underlying uncensored normal distribution (`mu`),
suitably adjusted with the cumulative distribution function and probability density function of the
uncensored distribution. For the three distributions in `Y` this yields:

```{r}
mean(Y)
```

In the first two the effect of censoring is rather small, while in the third distribution it is more
pronounced (due to the higher variance).

For evaluating the usual `d`/`p`/`q`/`r` functions in a standardized way, taking care of dimensions
and naming etc., the `distributions3` package provides the auxiliary function `apply_dpqr()`:


```{r}
cdf.Tobit <- function(d, x, drop = TRUE, elementwise = NULL, lower.tail = TRUE, log.p = FALSE, ...) {
  FUN <- function(at, d) {
    p <- pnorm(at, mean = d$mu, sd = d$sigma, lower.tail = lower.tail, log.p = log.p)
    p[rep_len(at, length(p)) < 0] <- if(lower.tail) {
      if(log.p) -Inf else 0
    } else {
      if(log.p) 0 else 1
    }
    p
  }
  apply_dpqr(d = d, FUN = FUN, at = x, type = "probability", elementwise = elementwise, drop = drop)
}
```

This can be used to evaluate all distribution functions at the same argument (returning a vector by default):

```{r}
cdf(Y, 0)
```

Or to evaluate all distribution functions at several arguments (returning a matrix by default):

```{r}
cdf(Y, c(0, 5))
```

Finally, if the distribution and the argument have the same length, the default is to do the evaluating
elementwise (always returning a vector):

```{r}
cdf(Y, 2:0)
cdf(Y, 2:0, elementwise = TRUE)
```

But the evaluation can also be forced to be carried out for each combination of distribution and argument (always returning a matrix):

```{r}
cdf(Y, 2:0, elementwise = FALSE)
```

These two additional methods are sufficient for the following illustrations. More details are provided below at the
end of this section.


## Adding a new model

When setting up the procasting infrastructure, the heavy lifting is done with the creation of a suitable
distribution class and methods. Interfacing a new model requires only a new `prodist()` method for setting
up the probability distribution object based on a fitted model (and potentially a `newdata` set). The idea
is to extract or predict the distribution parameters (`mu` and `sigma` in case of the Tobit distribution)
and subsequently call the distribution class constructor (preserving observation names, if any).

```{r}
prodist.crch <- function(object, newdata = NULL, na.action = na.pass, ...) {
  par <- predict(object, newdata = newdata, na.action = na.action, type = "parameter", ...)
  Tobit(mu = setNames(par$location, rownames(par)), sigma = par$scale)
}
```

To illustrate how this can be used in practice we fit a heteroscedastic Tobit model using
the `crch()` function to a precipiation dataset from Innsbruck, Austria. As precipitation
is often zero and never can become negative, a Tobit model is typically a good starting point
for probabilistic modeling and forecasting. (See FIXME for more details.)

The data can be preprocessed in the following way:

```{r}
data("RainIbk", package = "crch")
RainIbk <- sqrt(RainIbk)
RainIbk$ensmean <- apply(RainIbk[, grep('^rainfc', names(RainIbk))], 1, mean)
RainIbk$enssd   <- apply(RainIbk[, grep('^rainfc', names(RainIbk))], 1, sd)
RainIbk <- subset(RainIbk, enssd > 0)
```

And the model is fitted via:

```{r}
library("crch")
m <- crch(rain ~ ensmean | log(enssd), data = RainIbk, left = 0)
```

Then, the predicted Tobit distributions for the first two observations can be obtained with
the `prodist()` method obtained above:

```{r}
RainIbk2 <- head(RainIbk, 2)
prodist(m, newdata = RainIbk2)
```

Then, the `mean()` or `cdf()` method could be applied to this distribution vector.
Alternatively, instead of calling these methods "by hand", we can use the convenience
`procast()` method to do so:

```{r}
procast(m, newdata = RainIbk2, type = "mean")
procast(m, newdata = RainIbk2, type = "cdf", at = 0)
```

Instead of `"mean"` and `"cdf"`, we could also use the aliases `"response"` and `"probability"`,
respectively.

Similarly, other functions such as `proscore()`, `qresiduals()`, `pithist()`, or `rootogram()`
can be used once all of the methods from the table above are defined.


## Further details

Most of the remaining methods for the distribution objects follow the same structure as
the `mean()` and `cdf()` method above, respectively. However, the `random()` and the
`support()` methods are slightly different. Hence, these are also briefly illustrated 
in the following.

Drawing `random()` samples also uses `apply_dpqr()` with the argument `n` which is assured
to always be a positive integer.

```{r}
random.Tobit <- function(x, n = 1L, drop = TRUE, ...) {
  n <- make_positive_integer(n)
  if (n == 0L) {
    return(numeric(0L))
  }
  FUN <- function(at, d) {
    y <- rnorm(n = at, mean = d$mu, sd = d$sigma)
    y[y < 0] <- 0
    y
  }
  apply_dpqr(d = x, FUN = FUN, at = n, type = "random", drop = drop)
}
```

If `drop = TRUE` (the default), then one random sample for each distribution yields a vector,
while several random samples for each distribution yields a matrix:

```{r}
random(Y, 1)
random(Y, 3)
```

The `support()` method should return a matrix of `"min"` and `"max"` for the
distribution. The `make_support()` function helps to set the right names and
dimension.

```{r}
support.Tobit <- function(d, drop = TRUE, ...) {
  min <- rep(0, length(d))
  max <- rep(Inf, length(d))
  make_support(min, max, d, drop = drop)
}
support(Y)
```


## References

