---
title: "topmodels: Infrastructure for Inference and Forecasting in Probabilistic Models"
author: "Moritz Lang, Reto Stauffer, Achim Zeileis"
format: 
  html:
    html-math-method: mathjax
    toc: true
    number-sections: true
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{topmodels}
  %\VignetteEngine{quarto::html}
  %\VignetteDepends{topmodels,crch,ggplot2}
  %\VignetteKeywords{FIXME}
  %\VignettePackage{topmodels}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("topmodels")
options(digits = 4)
set.seed(123)
```

## Overview

Probabilistic predictions have been receiving increasing interest in various
application fields over the last decades due to necessary functional risk management and
strategy. Consequently, there is an increasing demand for appropriate probabilistic models
and corresponding evaluations of the goodness of fit. Besides proper probabilistic
scores [@Gneiting+Raftery:2007], which evaluate not only the expectation but the entire
predictive distribution, graphical assessment methods are particularly advantageous
to diagnose possible model misspecification problems.

Probabilistic predictions are often based on distributional regression models,
for which a wide range of different packages is readily available: from basic
models like `lm()` and `glm()` in base R (which can be interpreted as probabilistic
models and not just mean regression models), over general packages for distributional regression like
[`gamlss`](https://CRAN.R-project.org/package=gamlss) [@Stasinopoulos+Rigby:2007] or
[`bamlss`](https://CRAN.R-project.org/package=bamlss) [@Umlauf+Klein+Zeileis:2018, @Umlauf+Klein+Simon:2021]
to more specific packages for certain purposes. Examples for the latter include
[`pscl`](https://CRAN.R-project.org/package=pscl) or [`countreg`](https://CRAN.R-project.org/package=countreg)
[@Zeileis+Kleiber+Jackman:2008] for count regression,
[`crch`](https://CRAN.R-project.org/package=crch) [@Messner+Mayr+Zeileis:2016]
for certain censored regression models, or
[`betareg`](https://CRAN.R-project.org/package=betareg) [@Cribari-Neto+Zeileis:2010]
for beta regression, among many others.
However, there is no unified and object-oriented approach available for all these
different models/packages that allows to compute predictive distributions, probabilities, and
quantiles. Therefore, routines to evaluate probabilistic models either graphically
or via scoring rules are not always available or may be specific to certain packages.
An easy-to-use unified infrastructure for graphically
assessing and comparing different probabilistic models is not available, yet.

The
[**topmodels**](https://R-Forge.R-project.org/projects/topmodels/)
packages is designed to fill this gap and provide such an unifiying infrastructure to obtain
predictions of probabilities, densities, etc. for
probabilistic models. The unifying prediction infrastructure is the basis for numerous
graphical evaluation tools, such as rootograms [@Kleiber+Zeileis:2016], PIT
histograms [@Gneiting+Balabdaoui+Raftery:2007], reliagrams [reliability
diagrams, @Wilks:2011], randomized quantile Q-Q plots [@Dunn+Smyth:1996], and
worm plots [@Buuren+Fredriks:2001]. 

To be able to use the object-oriented framework of
[**topmodels**](https://R-Forge.R-project.org/projects/topmodels/pkg/topmodels/),
solely a `procast()` method must exist for the model class of interest.
Currently the package provides generic `procast` methods for the model classes
`lm`, `glm`, [`crch`](https://CRAN.R-project.org/package=crch)
[@Messner+Mayr+Zeileis:2016], and
[`disttree`](https://R-Forge.R-project.org/projects/partykit/pkg/disttree/)
[@Schlosser+Hothorn+Stauffer+Zeileis:2019].


## Installation

For the package
[**topmodels**](https://R-Forge.R-project.org/projects/topmodels/pkg/topmodels/)
so far only a development version is available, which is hosted on R-Forge at
<https://R-Forge.R-project.org/projects/topmodels/pkg/topmodels/> in a
Subversion (SVN) repository. The package can be installed via

```{r installation-cran, eval=FALSE}
install.packages("topmodels", repos = "https://R-Forge.R-project.org")
```

or via 

```{r installation-remotes, eval=FALSE}
remotes::install_svn("svn://R-Forge.R-project.org/svnroot/topmodels/pkg/topmodels")
```

where a specific revision can be installed by setting the optional argument `revision`.


## Usage

The package
[**topmodels**](https://R-Forge.R-project.org/R/?group_id=2305)
provides various routines to easily graphically assess and compare different
probabilistic models and model types using `ggplot2` [@Wickham:2016] and base
*R* graphics:

```{r example-model}
library("topmodels")
m <- lm(dist ~ speed, data = cars)
```

```{r example-rootogram}
rootogram(m)
```

```{r example-pithist}
pithist(m)
```

```{r example-qqrplot}
qqrplot(m)
```

```{r example-wormplot}
wormplot(m)
```

```{r example-reliagram}
reliagram(m)
```

## References


