---
title: "topmodels: Infrastructure for Inference and Forecasting in Probabilistic Models"
author: "Moritz Lang, Reto Stauffer, Achim Zeileis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{topmodels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{topmodels,crch,ggplot2}
  %\VignetteKeywords{FIXME}
  %\VignettePackage{topmodels}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("topmodels")
options(digits = 4)
set.seed(123)
```

## Overview

As a consequence of the growing importance of probabilistic predictions in
various application fields due to a necessary functional risk management and
strategy, there is an increasing demand for appropriate probabilistic model
evaluation. Besides proper probabilistic scores [@Gneiting+Raftery:2007], which
evaluate not only the expectation but the entire predictive distribution,
graphical assessment methods are particularly advantageous to diagnose possible
model misspecification problems.

Probabilistic predictions are often based on distributional regression models,
whereby the computation of predictive distributions, probabilities, and
quantiles is generally dependent on the *R* package being used. Therefore,
routines to graphically evaluate probabilistic models are not always available
and if so then only for specific types of models and distributions provided by
the corresponding package. An easy to use unified infrastructure to graphical
assess and compare different probabilistic model types does not yet exist.

The
[**topmodels**](https://r-forge.r-project.org/projects/topmodels/pkg/topmodels/)
packages is designed to provide such an unifiying infrastructure to gain
predictions of probabilities, densities, scores, and Hessian values for
probabilistic models. The unifying prediction method offers the user numerous
graphical evaluation tools, such as rootograms [@Kleiber+Zeileis:2016], PIT
histograms [@Gneiting+Balabdaoui+Raftery:2007], reliagrams [reliability
diagrams, @Wilks:2011], randomized quantile Q-Q plots [@Dunn+Smyth:1996], and
worm plots [@Buuren+Fredriks:2001]. 

To be able to use the object-oriented framework of
[**topmodels**](https://r-forge.r-project.org/projects/topmodels/pkg/topmodels/),
solely a `procast()` method must exist for the model class of interest.
Currently the package provides generic `procast` methods for the model classes
`lm`, `glm`, [`crch`](https://CRAN.R-project.org/package=crch)
[@Messner+Mayr+Zeileis:2016], and
[`disttree`](https://r-forge.r-project.org/projects/partykit/pkg/disttree/)
[@Schlosser+Hothorn+Stauffer+Zeileis:2019].


## Installation

For the package
[**topmodels**](https://r-forge.r-project.org/projects/topmodels/pkg/topmodels/)
so far only a development version is available, which is hosted on R-Forge at
<https://r-forge.r-project.org/projects/topmodels/pkg/topmodels/> in a
Subversion (SVN) repository. The package can be installed via

```{r installation-cran, eval=FALSE}
install.packages("topmodels", repos="http://R-Forge.R-project.org")
```

or via 

```{r installation-remotes, eval=FALSE}
remotes::install_svn("svn://r-forge.r-project.org/svnroot/topmodels/pkg/topmodels")
```

where a specific revision can be installed by setting the optional argument `revision`.


## Usage

The package
[**topmodels**](https://r-forge.r-project.org/projects/topmodels/pkg/topmodels/)
provides various routines to easily graphically assess and compare different
probabilistic models and model types using `ggplot2` [@Wickham:2016] and base
*R* graphics:

```{r example-model}
library("topmodels")
m <- lm(dist ~ speed, data = cars)
```

```{r example-rootogram}
rootogram(m)
```

```{r example-pithist}
pithist(m)
```

```{r example-qqrplot}
qqrplot(m)
```

```{r example-wormplot}
#```{r example-reliagram}
# FIXME: (ML) Produces error due to eval (something w/ enivornments not properly set?!)
#reliagram(m)
#````
wormplot(m)
```

## References


